---
title: "SF Final Assignment MÓF"
author: "Group 1"
date: "4/28/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(forecast)
```




1. First we read in the public transportation data.

```{r}
data <- read.csv("data/publicTransport_part.csv")
```

2. We must now create a time series object with the data.

```{r}
#data$DATE <- gsub("Mar", "03", data$DATE)
#data$DATETIME <- paste(data$DATE, data$TIME)
```




For our frequency, we are faced with a problem. The transport system starts at 6:30 in the morning and finishes at 22:00 in the evening. This means it does not operate on a 24 hour basis, and therefore we cannot use 96 (4 x 24) as our frequency.

In order to calculate the frequency we use the table function to return the number of unique values there are in the date column. This will tell us how many 15-minute timestamps there are per day, and thus the frequency

```{r}
table(data$DATE)
```
We can see that there is a frequency of 63, which we shall use. 

```{r}
transport.ts <- ts(data$DEMAND,
                   start = 6:30,
                   frequency = 63*7)

head(transport.ts)
tail(transport.ts)
```
```{r}
plot(transport.ts,
     main = "Public Transport usage over Time",
     ylab = "Usage (No. of passengers)",
     xlab = "Time (Date in March)",
     xaxt = "n")
axis(1,
     at = c(6.0:9.0),
     labels =c("Tue 1st", "Tue 8th", "Tue 15th", "Tue 22nd"))
grid(nx = NA, ny = NULL,
     lty = 2,
     col = "light gray",
     lwd = 2)
```
When we look at the data, we can notice that there isn't much of a trend. There is definitely seasonality however. We can see weekly seasonality - there is a spike in demand on each of the first five days, significant reduced demand on Saturday and even more reduced demand on Sunday. This occurs each week. 

```{r}
ggAcf(transport.ts)
ggsubseriesplot(transport.ts)
```

Preliminary pre processing:
If we are to do an 80/20 training/testing split, we can calculate the amount of data points needed per set by multiplying the total number of each by 0.8 and 0.2 respectively. 

```{r}
print(1323*0.8)
print(1323*0.2)
```


```{r}

```


```{r}
valid.n <- 265
training.n <- length(transport.ts) - valid.n
prediction.n <- 63*3

print(valid.n)
print(training.n)
print(prediction.n)
```

Creating training/validation split.

```{r}
train.ts <- window(transport.ts,
                   start = c(6.0, 1),
                   end = c(6.1, training.n))


valid.ts <- window(transport.ts,
                   start = c(6.1, training.n+0.1),
                   end = c(6.1, training.n+valid.n))

head(train.ts)
tail(train.ts)

head(valid.ts)
tail(valid.ts)
```
Remaking the testing and validation set.

```{r}
print(17*63)
```


Mean Prediction

```{r}
plot(meanf(train.ts, valid.n))
lines(valid.ts,
      col = "blue")
```

Naive Prediction

```{r}
naive.pred <- naive(train.ts,
                    h = valid.n)
plot(transport.ts,
     ylab = "Demand",
     xlab = "Time")
lines(valid.ts,
      col = "blue")
lines(naive.pred$mean,
      lty = 3,
      lwd = 2,
      col = "red")
```
```{r}
accuracy(naive.pred, valid.ts)
```


Drift prediction

```{r}
plot(rwf(train.ts, valid.n, drift = TRUE))
lines(valid.ts,
      col = "blue")
```

Seasonal Naive prediction

```{r}
snaive.pred <- snaive(train.ts,
                      h = valid.n)

plot(transport.ts,
     ylab = "Demand",
     xlab = "Time")
lines(valid.ts,
      col = "blue")
lines(snaive.pred$mean,
      lty = 3,
      lwd = 2,
      col = "red")
```

```{r}
accuracy(snaive.pred, valid.ts)
checkresiduals(snaive.pred)
```


TSLM with linear trend component

```{r}
train.lm <- tslm(train.ts~trend)

train.lm.pred <- forecast(train.lm, 
                          h = valid.n,
                          level = 1)

plot(train.lm.pred,
     ylab = "Demand",
     xlab = "Time",
     bty = "l",
     xaxt = "n",
     main = "",
     flty = 2)
lines(train.lm$fitted, lwd = 2)
lines(valid.ts)
```


```{r}
accuracy(train.lm.pred, valid.ts)
checkresiduals(train.lm)
```


TSLM with seasonal component

```{r}
train.lm.se <- tslm(train.ts ~ season)
train.lm.se.fc <- forecast(train.lm.se,
                           h = valid.n)

plot(train.ts,
     xlab = "Time",
     ylab = "Demand",
     bty = "l")
lines(train.lm.se$fitted, 
      lwd = 2,
      lty = 5,
      col = "red")
```

```{r}
accuracy(train.lm.se.fc, transport.ts)
checkresiduals(train.lm.se)
```


```{r}

```

Linear forecast model with trend and season

This model may not be necessary, because, as previously discussed there isn't much evidence of trend within the time series. Seasonality is already accounted for in the previous model.

```{r}
train.lm.tse <- tslm(train.ts ~ trend + season)
summary(train.lm.tse)

train.lm.tse.fc <- forecast(train.lm.tse,
                            h = valid.n)

plot(transport.ts,
     xlab = "Time",
     ylab = "Demand",
     bty = "l")
lines(train.lm.tse$fitted, 
      lwd = 1,
      lty = 2,
      col = "red")
lines(train.lm.tse.fc$fitted,
      col = "blue")
```

```{r}
accuracy(train.lm.tse.fc, transport.ts)
checkresiduals(train.lm.tse)
```



Advanced Exponential Smoothing forecast (ETS)

```{r}
train.ets <- ets(train.ts)
ets.fc <- forecast(train.ets, h = valid.n)

plot(transport.ts)
lines(train.ets$fitted,
      col = "green")
lines(valid.ts,
      col = "red")

```

```{r}
accuracy(ets.fc, valid.ts)
checkresiduals(ets.fc)
```


STLF Forecast

```{r}
train.stlf <- stlf(train.ts)
ets.stlf <- forecast(train.ets, h = valid.n)
```

```{r}
accuracy(ets.stlf, valid.ts)
checkresiduals(ets.stlf)
```

```{r}
plot(transport.ts)
lines(train.stlf$fitted,
      col = "green")
lines(valid.ts,
      col = "red")
```

Arima modelling

```{r}
train.ARIMA <- auto.arima(train.ts)
```

```{r}
train.ARIMA
summary(train.ARIMA)
```

```{r}
train.ARIMA.pred <- forecast(train.ARIMA,
                             h = valid.n)
```


```{r}
accuracy(train.ARIMA.pred, valid.ts)
checkresiduals(train.ARIMA.pred)
```

```{r}
plot(train.ARIMA.pred,
     ylab = "Demand",
     xlab = "Time")
lines(train.ARIMA.pred$fitted)
lines(valid.ts, col = "red")
```

```{r}
plot(transport.ts)
lines(train.ARIMA$fitted,
      col = "red")
```

```{r}
head(transport.ts)
head(train.ts)
```

```{r}
table(data$DEMAND)
```


Attempting to run a Prophet model on the time series data

```{r}
#install.packages('prophet')
```

```{r}
library(prophet)
library(tidyverse)
```

```{r}
#Prophet preprocessing

prophet.df <- data
prophet.df$ds <- paste(prophet.df$DATE, prophet.df$TIME)
str(prophet.df$ds)

prophet.df$ds <- str_replace_all(prophet.df$ds, "Mar", "03")

library(lubridate)
prophet.df$ds <- as.POSIXct(prophet.df$ds,
                            format = "%d-%m-%y %H:%M")
str(prophet.df$ds)
colnames(prophet.df)[3] = "y"

prophet.df <- prophet.df[-c(1,2)]
```

```{r}
p.model <- prophet(prophet.df)
p.future <- make_future_dataframe(p.model, periods = valid.n, freq = 96*7)
tail(p.future)
forecast.p <- predict(p.model, p.future)

tail(forecast.p)

dyplot.prophet(p.model, forecast.p)
```

```{r}
p.cv <- cross_validation(p.model,
                         initial = training.n,
                         period = 63,
                         horizon = valid.n,
                         units = "15 mins")
```




Final forecast using Time series Linear model with seasonality

```{r}
transport.lm.se <- tslm(transport.ts ~ season)
final.fc <- forecast(transport.lm.se,
                     h = prediction.n)
```

```{r}
plot(transport.ts,
     main = "Public Transport usage over Time (with forecast)",
     ylab = "Usage (No. of passengers)",
     xlab = "Time (Date in March)",
     xlim = c(6.0, 9.5),
     xaxt = "n",
     bty = "l")
grid(nx = NA, ny = NULL,
     lty = 2,
     col = "light gray",
     lwd = 2)
lines(train.ts,
      col = "black")
lines(train.lm.se$fitted, 
      lwd = 1,
      lty = 2,
      col = "red")
lines(train.lm.se.fc$mean,
      lwd = 1,
      lty = 2,
      col = "green")
lines(final.fc$mean, 
      lwd = 1,
      col = "blue")
axis(1,
     at = c(6.0:9.5),
     labels =c("Tue 1st", "Tue 8th", "Tue 15th", "Tue 22nd"))
axis(1,
     at = 9.45,
     labels = "25th")
legend(9.02, 140,
       legend = c("Time series", "Training", "Validation", "Forecast"),
       col = c("black","red","green","blue"),
       lty = c(1, 2, 2, 2),
       cex = 0.6)

```

```{r}
plot(final.fc$mean,
     main = "Public Transport Usage Forecast",
     ylab = "Usage (No. of passengers)",
     xlab = "Time (Date in March)",
     lwd = 2,
     col = "blue",
     xaxp = c(0, 5, 4))
grid(nx = NULL, ny = NULL,
     lty = 2,
     col = "light gray",
     lwd = 2)
axis(1,
     at = c(9.01, 9.16, 9.30,9.44),
     labels =c("22nd", "23rd", "24th", "25th"))

```

```{r}
accuracy(transport.lm.se$fitted.values, transport.ts)
checkresiduals(transport.lm.se)
```